---
title: "Population Profile of Sri Lanka"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
css: custom.css
---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(tsibble)
library(lubridate)
library(htmltools)
library(readr)
library(readxl)
library(ggplot2)
library(plotly)
library(knitr)
library(RColorBrewer)
library(scales)
library(patchwork)
library(ceylon)
library(sf)

# datasets
table01_page1 <- read_csv("table01_page1.csv")
population_density_district <- read_csv("population_density_district.csv")
province_district <- read_excel("province_district.xlsx")
mid_year_pop_district <- read.csv("mid_year_pop_district.csv")
srilanka_gender <- read.csv("srilanka_gender.csv")
cleaned_age_groups_data <- read.csv("cleaned_age_groups_data.csv")
district_pop <- read.csv("district_pop.csv")


# Data preprocess
# dataset 01
table01_page1_ts <- table01_page1 %>%
  as_tsibble(index = `census year`)

table01_page1_ts <- table01_page1_ts %>%
  mutate(`Average annual growth rate` = as.numeric(`Average annual growth rate`))

# Dataset 02
sl_density <- data.frame(
  Year = c(1981, 2001, 2012, 2024),
  Density = c(230, 300, 325, 349)
)

sl_density <- sl_density %>%
  as_tsibble(index = Year)

# Vector of province names
provinces <- c("Sri Lanka", "Western", "Central", "Southern", "Northern", 
               "Eastern", "North Western", "North Central", "Uva", "Sabaragamuwa")

# Extract province rows
sl_province_df <- province_district %>%
  filter(`Province and district` %in% provinces)

# Extract district rows (everything else)
district_df <- province_district %>%
  filter(!(`Province and district` %in% provinces))
```


```{r}
htmltools::tags$div(
  id = "top-right-logo",
  tags$img(src = "sri_lanka_logo.png", alt = "Sri Lanka Logo")
)
```



# National Overview

## column { data-width = "450"}

###
  
```{r}
valueBox(
  value = max(table01_page1_ts$`census date`),
  caption = "Most Recent Census Date",
  icon = "fa-calendar",
  color = "#510400"
)
```

###

```{r}
# Create a value box showing the latest population
valueBox(
  value = format(max(table01_page1_ts$`Enumerated Population`), big.mark = ","),
  caption = "Population",
  icon = "fa-users",
  color = "darkgreen"
)
```

###

```{r}
sl_population <- sl_province_df %>%
  filter(`Province and district` == "Sri Lanka") %>%
  pull(`usual residence population`)

valueBox(
  value = format(sl_population, big.mark = ","),
  caption = "Usual Residence Population",
  icon = "fa-home",
  color = "darkred"
)
```

###

```{r}
latest_density <- sl_density %>%
  filter(Year == max(Year)) %>%
  pull(Density)

valueBox(
  value = format(latest_density, big.mark = ","),
  caption = "Population Density (persons per square km)",
  icon = "fa-city",
  color = "goldenrod"
)
```

###

```{r, include=FALSE}
SL_mid_year_pop_2024 <- data.frame(
  Gender = c("Female", "Male", "Total"),
  Population = c(11305, 10611, 21916)
)
```

```{r}
valueBox(
  value = comma( SL_mid_year_pop_2024$Population[SL_mid_year_pop_2024$Gender == "Female"]*1000),
  caption = "Female Mid Year Population - 2024",
  icon = "fa-female",
  color = "#E377C4"
)
```

###

```{r}
# Value box for Male population
valueBox(
  value = comma( SL_mid_year_pop_2024$Population[SL_mid_year_pop_2024$Gender == "Male"]*1000),
  caption = "Male Mid Year Population - 2024",
  icon = "fa-male",
  color = "#1F77B4"
)
```

## column { data-width = "400"}

###

```{r}
p1 <- ggplot(table01_page1_ts, aes(x = `census year`, y = `Enumerated Population`)) +
  geom_line(color = "darkgreen", size = 1) +
  geom_point(color = "darkgreen", size = 2) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    x = "Census Year",
    y = "Population"
  ) +
  theme_minimal()

ggplotly(p1)
```

### Midyear Population by Gender Over Time

```{r}
srilanka_gender_ts <- srilanka_gender %>%
   mutate(Gender = factor(Gender, levels = c("Female", "Male", "Total"))) %>%
   as_tsibble(index = Year, key = Gender)
 
 gender_colors <- c(
   "Female" = "#E377C2",  # rose pink
   "Male"   = "#1F77B4",  # blue
   "Total"  = "darkgreen"   # green
 )
 
p2 <- ggplot(srilanka_gender_ts, aes(x = Year, y = Population, color = Gender)) +
   geom_line(size = 1.2) +
   geom_point(size = 2) +
   labs(y = "Population (thousands)") +
   scale_color_manual(values = gender_colors) +
   theme_minimal()
 
ggplotly(p2)
```

## column { data-width = "400"}

###

```{r}
p3 <- ggplot(table01_page1_ts, aes(x = `census year`, y = `Average annual growth rate`)) +
  geom_line(color = "orange", size = 1) +      
  geom_point(color = "orange", size = 2) +     
  labs(
    x = "Census Year",
    y = "Average Annual Growth Rate (%)"
  ) +
  theme_minimal() 

ggplotly(p3)
```

### Mid year Population by Age Group, 2024

```{r}
p4 <- cleaned_age_groups_data %>%
  filter(Year == 2024, Gender == "Total") %>%
  mutate(Age_Group = factor(Age_Group, levels = c(
    "0-4", "5-9", "10-14", "15-19", "20-24", "25-29", "30-34", "35-39", 
    "40-44", "45-49", "50-54", "55-59", "60-64", "65-69", "70-74", "75-79", "80+"))) %>%
  ggplot(aes(x = Age_Group, y = Population)) +
  geom_col(fill = "darkblue") +
  labs(
    x = "Age Group",
    y = "Population (thousands)"
  ) +
  coord_flip() +
  theme_minimal()

ggplotly(p4)
```



# Provincial Overview

## column { data-width = "400"}  

### Provincial Population, 2024

```{r, include=FALSE}
# Provinces list
provinces <- c(
  "Western", "Central", "Southern", "Northern", 
  "Eastern", "North Western", "North Central", 
  "Uva", "Sabaragamuwa"
)

# Extract only province rows
province_pop <- province_district %>%
  filter(`Province and district` %in% provinces) %>%
  rename(Province = `Province and district`)

province_pop <- province_pop %>%
  rename(PROVINCE = Province)

province_pop <- province_pop %>%
  mutate(PROVINCE = toupper(PROVINCE))

province <- ceylon::province

province_map_data <- province %>%
  left_join(province_pop, by = "PROVINCE") 
```

```{r}
p5 <- ggplot(province_map_data) +
  geom_sf(aes(
    fill = `total population`,
    text = paste0("Province: ", PROVINCE, "<br>",
                  "Total Population: ", `total population`)
  )) +
  scale_fill_distiller(
    palette = "Greens",  
    name = "Total Population",
    direction = 1,
    labels = label_comma() 
  ) +
  theme_minimal()

ggplotly(p5, tooltip = "text")
```

## column { data-width = "400"}

### Roofless Percentage, 2024

```{r}
p6 <- ggplot(province_map_data) +
    geom_sf(aes(fill = percentage,
                text = paste0("Province: ", PROVINCE, "<br>",
                  "Percentage: ", `percentage`)
                )) +
 scale_fill_distiller(
    palette = "Reds",  
    name = "Percentage",
    direction = 1
  ) +
theme_minimal()

ggplotly(p6, tooltip = "text")
```

## column { data-width = "400"}

### Genderwise Mid Year Population, 2024

```{r, inlcude = FALSE}
# Mapping: District → Province
district_province_map <- tribble(
  ~Region,         ~Province,
  "Colombo",       "Western",
  "Gampaha",       "Western",
  "Kalutara",      "Western",
  "Kandy",         "Central",
  "Matale",        "Central",
  "Nuwara-eliya",  "Central",
  "Galle",         "Southern",
  "Matara",        "Southern",
  "Hambantota",    "Southern",
  "Jaffna",        "Northern",
  "Kilinochchi",   "Northern",
  "Mannar",        "Northern",
  "Vavuniya",      "Northern",
  "Mullaitivu",    "Northern",
  "Batticaloa",    "Eastern",
  "Ampara",        "Eastern",
  "Trincomalee",   "Eastern",
  "Kurunegala",    "North Western",
  "Puttalam",      "North Western",
  "Anuradhapura",  "North Central",
  "Polonnaruwa",   "North Central",
  "Badulla",       "Uva",
  "Monaragala",    "Uva",
  "Ratnapura",     "Sabaragamuwa",
  "Kegalle",       "Sabaragamuwa"
)

# Summarise: Province + Year + Gender totals
province_pop_gender <- mid_year_pop_district %>%
  left_join(district_province_map, by = "Region") %>%
  group_by(Province, Year, Gender) %>%
  summarise(
    total_population = sum(Population, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
p7 <- province_pop_gender %>%
  filter(Year == 2024, Gender != "Total") %>%
  mutate(Province = reorder(Province, total_population, sum)) %>%
  ggplot(aes(x = Province, y = total_population, fill = Gender)) +
  geom_col(position = "dodge") +
  scale_fill_manual(
    values = c("Male" = "#1F77B4", "Female" = "#E377C4")
  ) +
  labs(x = "Province", y = "Population (thousands)") +
  theme_minimal() +
  coord_flip()

ggplotly(p7)
```

### Mid Year Population Trends

```{r}
p8 <- province_pop_gender %>%
  filter(Gender == "Total") %>%
  ggplot(aes(x = Year, y = total_population, color = Province)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    x = "Year",
    y = "Population (thousands)",
    color = "Province"
  ) +
  theme_minimal()

ggplotly(p8)
```

# District level drill down

## column { data-width = "400"}

### District Population, 2024

```{r, include=FALSE}
district_pop <- district_pop %>%
  mutate(District = toupper(District))

district <- ceylon::district

district_map_pop <- district %>%
  left_join(district_pop, by = c("DISTRICT" = "District"))
```

```{r}
p9 <- ggplot(district_map_pop) +
  geom_sf(aes(fill = total_population,
              text = paste0("District: ", DISTRICT,
                            "<br>",
                            "Total Population: ", comma(total_population)))) +
  scale_fill_distiller(palette = "Greens",
                       direction = 1,
                       name = "Total Population",
                       labels = comma) +
  theme_minimal()

ggplotly(p9, tooltip = "text")
```

## column { data-width = "400"}

### Population Density, 2024

```{r}
p10 <- ggplot(district_map_pop) +
  geom_sf(aes(fill = population_density,
              text = paste0("District: ", DISTRICT,
                            "<br>",
                            "Population Density: ", comma(population_density)))) +
  scale_fill_distiller(palette = "Blues", direction = 1, 
                       name = "Population Density") +
  theme_minimal()

ggplotly(p10, tooltip = "text")
```

## column { data-width = "400"}

### Mid Year Population Over Time

```{r}
mid_year_pop_ts <- mid_year_pop_district %>%
  filter(Gender == "Total") %>%   # Keep only total population
  as_tsibble(
    key = Region,                  # Each district is a "key"
    index = Year                   # Time index column
  )

plot_ly(mid_year_pop_ts, x = ~Year, y = ~Population, color = ~Region,
        type = 'scatter', mode = 'lines+markers') %>%
  layout(
    xaxis = list(title = "Year"),
    yaxis = list(title = "Population (thousands)")
  )
```


